### 2024-07-10 cognitive-bias.qmd
20240710 - overconfidence

>>>>

In general, people tend to be overconfident in their judgments and predictions.
*Overconfidence bias* is the tendency for a person to have greater confidence in their abilities (including judgments and predictions) than is objectively warranted.
There are three general ways that overconfidence has been identified [@Moore2008]:

1. *overestimation* of one's actual performance
1. *overplacement* of one's performance relative to others
1. *overprecision* in one's beliefs/jugments/predictions

*Overestimation* involves believing that one will perform better than one actually performs.
Overestimation can be identified with a [calibration plot](#sec-calibration) of the predicted performance versus actual performance, where the person's predicted performance is systematically higher (in at least some cases) than their actual performance.
Overestimation corresponds to the "[overprediction](#sec-calibration)" form of [miscalibration](#sec-calibration).

*Overplacement* involves believing that one is better than others or will perform better than others, even when they do not.
For instance, it is a common finding that more than half of people believe they are "above average" (i.e., above the median), even though that is statistically impossible.
This calls to mind the fictitious Lake Wobegon in the radio show *A Praririe Home Companion*, "where all the women are strong, all the men are good-looking, and all the children are above average."

*Overprecision* involves expressing excessive certainty regarding the accuracy of one's beliefs/judgments/predictions.
For instance, if when a given weather forecaster says it will rain 80% of the time, it actually rains 30% of the time, the weather forecaster's predictions are overprecise.
Likewise, if the weather forecast says it will rain 10% of the time and it actually rains 30% of the time, the predictions are also overprecise because the forecaster is expressing stronger confidence than is warranted that it will not rain.
Overprecision can be identified with a [calibration plot](#sec-calibration) of the predicted probabilities versus the actual probabilities.
Overprecision corresponds to the "[overextremity](#sec-calibration)" form of [miscalibration](#sec-calibration).

A fantasy manager may be even more likely to exhibit overconfidence if they previously performed well or won their league, for which luck and random chance plays an important role.
Indeed, it is estimated that nearly half (~45%) of the variability in fantasy football performance is estimated to be luck [and around 55% due to skill; @Getty2018].
A manager who won their league in the prior season may believe they will perform better than they actually will (overestimation), will perform better than average (overplacement), and may hold excessive confidence regarding the accuracy of their predictions about which players will perform well or poorly (overprecision).
These various types of overconfidence may lead them to draft high-risk players based on gut feeling, neglecting statistical analysis and expert consensus.

### 2024-07-10 evaluating-prediction-accuracy.qmd
20240605 - getting started/load packages
20240608 - getting started
20240627 - sessionInfo() for reproducibility
20240627 - sessionInfo()
20240703 - conclusion section
20240710 - overconfidence

>>>>

## Getting Started {#sec-predictionAccuracyGettingStarted}

### Load Packages {#sec-predictionAccuracyLoadPackages}

```{r}

```

# Types of Accuracy {#sec-accuracyTypes}

## Discrimination {#sec-discrimination}

[Discuss options for discrete versus continuous data]

## Calibration [#sec-calibration]

[Discuss options for discrete versus continuous data]

The various types of miscalibration are depicted in @fig-miscalibration:

```{r}
#| label: fig-miscalibration
#| fig-cap: "Types of Miscalibration. From @Petersen2024a and @PetersenPrinciplesPsychAssessment."
#| code-fold: true

# Specify data
examplePredictions <- seq(from = 0, to = 1, by = .1)
exampleOutcomes <- c(0, .15, .3, .4, .45, .5, .55, .6, .7, .85, 1)

overPrediction <- c(0, .02, .05, .1, .15, .2, .3, .4, .5, .7, 1)
underPrediction <- c(0, .3, .5, .6, .7, .8, .85, .9, .95, .98, 1)
overExtremity <- c(0, .3, .38, .42, .47, .5, .53, .58, .62, .7, 1)
underExtremity <- c(0, .05, .08, .11, .2, .5, .8, .89, .92, .95, 1)

# Plot
par(
  mfrow = c(2,2),
  mar = c(5,4,1,1) + 0.1) #margins: bottom, left, top, right

plot(
  examplePredictions,
  overExtremity,
  xlim = c(0,1),
  ylim = c(0,1),
  main = "Overextremity",
  xlab = "Predicted Probability",
  ylab = "Observed Proportion",
  bty = "l",
  cex = 1.5,
  col = "#e41a1c",
  type = "o")

lines(
  c(0,1),
  c(0,1),
  lwd = 2,
  col = "#377eb8")

plot(
  examplePredictions,
  underExtremity,
  xlim = c(0,1),
  ylim = c(0,1),
  main = "Underextremity",
  xlab = "Predicted Probability",
  ylab = "Observed Proportion",
  bty = "l",
  cex = 1.5,
  col = "#e41a1c",
  type = "o")

lines(
  c(0,1),
  c(0,1),
  lwd = 2,
  col = "#377eb8")

plot(
  examplePredictions,
  overPrediction,
  xlim = c(0,1),
  ylim = c(0,1),
  main = "Overprediction",
  xlab = "Predicted Probability",
  ylab = "Observed Proportion",
  bty = "l",
  cex = 1.5,
  col = "#e41a1c",
  type = "o")

lines(
  c(0,1),
  c(0,1),
  lwd = 2,
  col = "#377eb8")

plot(
  examplePredictions,
  underPrediction,
  xlim = c(0,1),
  ylim = c(0,1),
  main = "Underprediction",
  xlab = "Predicted Probability",
  ylab = "Observed Proportion",
  bty = "l",
  cex = 1.5,
  col = "#e41a1c",
  type = "o")

lines(
  c(0,1),
  c(0,1),
  lwd = 2,
  col = "#377eb8")
```

## General Accuracy {#sec-generalAccuracy}

[Discuss options for discrete versus continuous data]

## Conclusion {#sec-predictionAccuracyConclusion}

::: {.content-visible when-format="html"}

## Session Info {#sec-predictionAccuracySessionInfo}

```{r}
sessionInfo()
```

:::
